Verifying Spatial Properties of Stencil Computations

### 2016-11-16 paper.tex
section 5 polishing / shrinking

To support the development of correct stencil computations, we propose
a simple, abstract specification language for the data access pattern
of stencils. The shape of the stencil computation in the
Laplace example is specifed in our language as:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a symmetrical pattern
(``centered'') to a depth of one in each direction in its first
dimension.  The shape of the Navier-Stokes example
is specified as:
%
\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
!= stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u

@@

\begin{definition}[Induction variable]
  An integer variable is an \emph{induction variable} if it is the control
  variable of a \texttt{for} loop incremented by $1$ per iteration. A variable
  is interpreted as an induction variable only within the scope of the loop
  body. Induction varaibles range over $i, j, k$ throughout.
%TODO: consider mentioning derived, non-based induction variables
% as we did before.
\end{definition}

\begin{definition}[Array subscripts and indices]
  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
  which reads from an array $a$ at an \emph{index} specified by a
  comma-separated sequence of integer expressions denoted $\bar{e}$ or
  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$
  is called \emph{relative} if the expression involves an induction
  variable.

  \emph{Absolute index} is a constant integer expression relative to the
  enclosing loop.
\end{definition}

@@

Now we informally describe what it means for a stencil kernel to be consistent
with a specification. This will will be formalised in \S~\ref{sec:semantics} and
algorithmically described in  and \S~\ref{sec:analysis}.
Specifications are associated to a stencil kernel and are declared for one or
more array variables $\bar{v}$ within that kernel.  A stencil specification
defines a \emph{region} associated to an array variable $a$ against which a
stencil kernel should be \emph{consistent}. The region defined by a
specification is an $n$-dimensional rectilinear shape in a discrete space
crossing the origin with respect to the induction variable. Consistency is then
a two way requirement. All array subscripts on $a$
contributing to the kernel should be consistent with the region. Conversely,
all parts of the region should have an index corresponding to that part.

@@

A \term{forward} region of depth $n$ specifies a continguous
region in dimension $d$ starting at the origin. This corresponds
to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
for some induction variable $i$. Similarly, a
\term{backward} region of depth $n$ corresponds to contiguous indices
from $i$ to $i - n$ and \term{centered}
of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
stencil specifies a neighbour index $i$. For example, the
following shows four specifications with four consistent stencil
kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d} respectively:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
!= stencil backward(depth=2, dim=1) :: b
e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
!= stencil centered(depth=1, dim=2) :: c
e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
!= stencil pointed(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
Not every dimension needs to be specified. In the above, the
specifications on line $1$, $3$, and $7$ leaves some dimensions unspecified.

@@

The region operators \term{+} and \term{*} provide ways to combine
regions disjunctively and conjuctively respectively.

@@

Region specifications can be assigned to region variables via
region declarations. For example, the last specification
 example above can be restated as:
%%
\begin{minted}{fortran}
!= region :: r1 = centered(depth=1, dim=1) * pointed(dim=2)
!= region :: r2 = centered(depth=1, dim=2) * pointed(dim=1)
!= region :: fivePoint = r1 + r2
!= stencil fivePoint :: a
\end{minted}
This is especially useful for common stencil patterns, such as five point
stencil, as the region can be defined once and reused many times later.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in Fig.~\ref{fig:syntax}).

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, in all of
the previous examples the \texttt{readOnce} modifier could be added,
\eg{}
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of \emph{linearity} in linear type
systems~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

Indexing schemes give an abstracts array subscripts.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours which are consistent with the specification.
The relationship between source language
terms and indexing terms
 is given precisely by the following function \textsf{schematic},
 which forms part of checking and inference (\S\ref{sec:analysis}).

\newcommand{\schemeOne}{\textsf{s}_I}
%\newcommand{\schemeOne}{\textsf{schema$^1_I$}}
\newcommand{\schemeN}{\textsf{schematic}_I}
%\newcommand{\schemeN}{\textsf{schema$^n_I$}}
\begin{definition}%[Schematic]
\label{def:schematic}
Let $\schemeOne$ be a partial function, parameterised by a set of
induction variables $I$, mapping expression syntax to
$\domainVal$ values:
%%
\begin{align*}
\schemeOne(e)
 & =  \begin{cases}
c & e \equiv i + c \;\; \wedge \;\; i \in I \\
-c & e \equiv i - c \;\; \wedge \;\; i \in I \\
\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
In the
first two cases, expressions that are a constant offset of an
induction variable (\ie{}, neighbourhood indices) are mapped to the
offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
expression is not defined in terms of an induction variable, it
is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
is undefined (returns $\bot$).

@@

Specifications have a theory of equality $\equiv$ as well
as approximation $<:$ which we define here.
For brevity we use abbreviations $\term{fwd}$, $\term{bwd}$, and
$\term{cen}$ for region constants here.

The equality relation $\equiv$ is defined over \emph{region}
syntax which gives the following axioms on regions:
%
\begin{itemize}
\item $\term{+}$ is idempotent, commutative, and associative;
\item $\term{*}$ is commutative, associative, and distributes over
$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)$. 
Distributivity is 
used later to give a normal form for specifications
akin to Disjunctive Normal Form (\S~\ref{sec:inf-step4}).

@@

\begin{figure}[t]
\vspace{-0.5em}
\begin{align*}
\hspace{-0.5em}
{\small{
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%%
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
%  Q}\trule{\textsc{trans}}
%\qquad
%\\[1.5em]
\vspace{-0.4em}
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
\end{array}}\trule{\textsc{over}} \\[-0.2em]
%\hspace{-0.5em}
%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
%      {R_1 \, \texttt{+} \, R_2 <:^r S_1 \, \texttt{+} \, S_2}
%\trule{\textsc{cong}\texttt{+}}
%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
%      {R_1 \, \texttt{*} \, R_2 <:^r S_1 \, \texttt{*} \, S_2}
%\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
%
\dfrac{R <:^r S}{\term{atLeast}  S <: \term{atLeast}
  R}\trule{\texttt{L}}
\;\;
\dfrac{R <:^r S}{\term{atMost}  R <: \term{atMost}
  S}\trule{\texttt{M}} \\[1.5em]
\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{grow}} \,
\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{shrink}} \,
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}}
 \\[1.5em]
%%
\dfrac{\mathsf{dims}(R) \cap \mathsf{dims}(S) = \emptyset}{R \ast S <: R}
\trule{\textsc{Gen$\ast$}}
\quad
%%
%\\[1.5em]
%\dfrac{R <: S}{\term{readOnce} \, R <: \term{readOnce} \,
%  S}\trule{\textsc{cong}\texttt{R}}
%\;\;
\;
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
%\dfrac{}{R <: \term{atMost} R}
%\quad
%\dfrac{}{R <: \term{atLeast} R}\\[1.5em]
\end{array}}}
\end{align*}
\caption{Inequations $<:^r$ on \textit{regions} and
$<:$ on \textit{specs} (omitting reflexivity, transitivity, and congruences). }
\label{fig:inequations}
\vspace{-0.8em}
\end{figure}

@@

\noindent
We outline here the procecures for checking the conformance
of source code against specifications (\S~\ref{subsec:checking})
and for infering specifications from code (\S~\ref{subsec:inference}).
Both rely on the analysis of source code, converting array indexing
expressions into sets of index schemes. We outline this analysis
first (\S~\ref{subsec:analysis}). Note that the analysis procedure
can be made arbitrarily more complex and wide-ranging independent
of the checking and inference procedures.

\subsection{Analysis of array-based code}
\label{subsec:analysis}

@@

\paragraph{Syntax helper functions}
The function \rhsExp{} maps source statements to a
set of expressions which occur in right-hand positions (\ie{}, not the
target of an assignment). The function \var{} maps expressions to a
set of the variables used in right-hand positions.

\paragraph{Analysis pre-requisites}
The analysis builds on standard program analyses:
%
\textbf{(1)} basic blocks (CFG);
\textbf{(2)} induction variables per basic block;
\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\textbf{(4)} type information per variable, where we use the predicate
\arrayTy{} to classify variables of array type.

\paragraph{Core analysis}
For each assignment statement \textit{assg} within a loop whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed mapping
array variables to \emph{multisets} of indexing schemes.
This is captured by the function $\textit{analyse}$:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textit{analyse}(a(\overline{e_1}) = e_2)
 := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(a')\}}
\end{array}
\end{align*}
%
The predicate $\neigh$ classifies neighbourhood indices, based on
the induction variables of the loop (implicit here).
Thus, we focus on assignments to an array subscript where the LHS
index $\overline{e_1}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a \emph{multiset} of indexing schemes for its subscripts, computed
with \textsf{schematic} (Def.~\ref{def:schematic}).

Note that \textsf{schematic} is undefined ($\bot$) if
$\bar{e}$ contains relative indices which are not neighbourhood
indices. The \textit{analysis} function feeds into the
\textit{modelise} function which filters out those arrays
which contain undefined indexing schemes and adds multiplicity
information:
%
\begin{align*}
& \mathit{modelise} (\textit{assg})
= a \mapsto
\begin{cases}
\bot & \bot \in M \\
\textsf{only}(|M|) & \forall u \in M . \#(M, u) = 1 \\
\textsf{mult}(|M|) & \forall u \in M . \#(M, u) = n
\end{cases} \\
& \quad \textit{where} \; M = \mathit{analyse}(\textit{assg}) (a)
\end{align*}
where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
multiset $M$ and $|M|$ denotes the set of elements in the multiset
$M$.

Thus, the analysis maps source code to index schemes, which is then
augmented with multiplicity information by $\mathit{modelise}$,
\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.

If the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets
then all RHS schemes are \emph{relativised} by
$\textsf{schematic}(\overline{e_1})$. For example,
if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.
\dnote{Somewhere we need a discussion about syntactic vs semantic
analysis...}

\begin{example}
\label{exm:checking}
We demonstrate the analysis and modelise procedures over the
following example program which computes the mean value
of a five-point stencil:
%%
\begin{minted}{fortran}
do i = 1, (n-1)
  do j = 1, (m-1)
    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
    b(i, j) = (a(i, j) + x + y) / 5.0
  end do
end do
\end{minted}
%
The \textit{analyse} function matches on
line 4, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textit{analyse} on line 4 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}^1, \vtwoh{1}{0}^1,
          \vtwoh{0}{-1}^1, \vtwoh{0}{1}^1, \vtwoh{0}{0}^1\}
\end{align*}
%
where multiplicities are denoted by superscripts. The \emph{modelise}
function then yields the following since all multiplicities in the model are one:
%%
\begin{align*}
\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
\end{align*}
%%
\end{example}

@@

In checking mode, the access pattern of stencil computations in the
source language is verified against any specification associated with
them in the code. The model of
\S~\ref{sec:semantics} guides this process.  Checking
proceeds by generating a model from a specification and generating
a model from the source code, comparing them for
\emph{consistency}. The notion of consistency is built out of index scheme
\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}).

\begin{definition}[Checking]
Let $a$ be an array variable and let
$S \in \textit{spec}$ be a specification associated to $a$
via \texttt{!= stencil $S$ :: a}. A stencil kernel
 $v(\overline{e_1}) = e_2$ in the source language
conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds,
defined:
\begin{align*}
& \textbf{check}_n(S, \textit{assg}, a) =
  \consSub{n}{\textit{modelise}(\textit{assg})(a)}{\interp{S}_n}
&
\end{align*}
Thus, checking proceeds by using \textit{modelise}
(\S~\ref{subsec:analyis}) to extract index schemes from source code
augmented with multiplicitly for the whole set of schemes.
This is compared with the model of the specification $\interp{S}_n$ by
the consistency predicate $\consAName$.
\end{definition}

>>>>

To support the development of correct stencil computations, we propose
a simple, abstract specification language for the data access pattern
of stencils. The shape of the
Laplace example stencil is specifed in our language as:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes \texttt{a} is accessed with a symmetrical pattern
(``centered'') to a depth of one in each direction in its first
dimension.  The shape of the Navier-Stokes example
is specified as:
%
\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
!= stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u

@@

\begin{definition}[Induction variable]
  An integer variable is an \emph{induction variable} if it is the control
  variable of a ``\texttt{for}'' loop incremented by $1$ per iteration. A variable
  is interpreted as an induction variable only within the scope of the loop
  body. Throuhgout, $i, j, k$ range over induction variables.
%TODO: consider mentioning derived, non-based induction variables
% as we did before.
\end{definition}

\begin{definition}[Array subscripts and indices]
  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
  which reads from an array $a$ at an \emph{index} specified by a
  comma-separated sequence of integer expressions denoted $\bar{e}$ or
  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$
  is called \emph{relative} if the expression involves an induction
  variable.
  An \emph{absolute index} is a constant integer expression relative to the
  enclosing loop.
\end{definition}

@@

Now we informally describe what it means for a stencil kernel to be consistent
with a specification. This will be formalised in \S~\ref{sec:semantics} and
algorithmically described in  and \S~\ref{sec:analysis}.
Specifications are associated to a stencil kernel and are declared for one or
more array variables $\bar{v}$ within that kernel.  A stencil specification
defines a \emph{region} associated to an array variable $a$ against which a
stencil kernel should be \emph{consistent}. The region defined by a
specification is an $n$-dimensional rectilinear shape in a discrete space
crossing the origin with respect to the induction variable. Consistency is then
a two way requirement. All array subscripts on $a$
contributing to the kernel should be consistent with the region. Conversely,
all parts of the region should have an index corresponding it.

@@

A \term{forward} region of depth $n$ specifies a continguous
region in dimension $d$ starting at the origin. This corresponds
to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
for some induction variable $i$. Similarly, a
\term{backward} region of depth $n$ corresponds to contiguous indices
from $i$ to $i - n$ and \term{centered}
of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
stencil specifies a neighbour index $i$. For example, the
following shows four specifications with four consistent stencil
kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d} respectively:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
!= stencil backward(depth=2, dim=1) :: b
e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
!= stencil centered(depth=1, dim=2) :: c
e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
!= stencil pointed(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
Not every dimension needs to be specified, \eg{},
specifications on lines $1$, $3$, and $7$ leave some dimensions unspecified.

@@

The region operators \term{+} and \term{*} respectively combine
regions disjunctively and conjuctively.

@@

Region specifications can be assigned to region variables via
region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common stencil patterns, such as the five-point
pattern, as the region can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in Fig.~\ref{fig:syntax}).

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of \emph{linearity} in linear type
systems~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

Indexing schemes give an abstract
representation of array subscripts in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours which are consistent with the specification.
The relationship between source language
terms and indexing terms
 is given precisely by the following function \textsf{schematic},
 which forms part of checking and inference (\S\ref{sec:analysis}).

\newcommand{\schemeOne}{\textsf{s}_I}
%\newcommand{\schemeOne}{\textsf{schema$^1_I$}}
\newcommand{\schemeN}{\textsf{schematic}_I}
%\newcommand{\schemeN}{\textsf{schema$^n_I$}}
\begin{definition}%[Schematic]
\label{def:schematic}
Let $\schemeOne$ be a partial function, parameterised by a set of
induction variables $I$, mapping expression syntax to
$\domainVal$ values:
%%
\begin{align*}
\schemeOne(e)
 & =  \begin{cases}
c & e \equiv i + c \;\; \wedge \;\; i \in I \\
-c & e \equiv i - c \;\; \wedge \;\; i \in I \\
\infty & \textit{$I \not\subseteq \textit{FV}(e)$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
where $\textit{FV}(e)$ is the set of free variables in $e$. In the
first two cases, expressions that are a constant offset of an
induction variable (\ie{}, neighbourhood indices) are mapped to the
offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
expression is not a neighbourhood index, but neither is it a relative
index (\ie{}, it is not defined in terms of an induction variable) then it
is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
is undefined (returns $\bot$).

@@

Specifications have a theory of equality $\equiv$ as well
as approximation $<:$ which we define here.
For brevity here we use abbreviations $\term{fwd}$, $\term{bwd}$, and
$\term{cen}$ for region constants.

The equality relation $\equiv$ is defined over \emph{region}
syntax which gives the following axioms:
%
\begin{itemize}
\item $\term{+}$ is idempotent, commutative, and associative;
\item $\term{*}$ is commutative, associative, and distributes over
$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)$.
Distributivity is
used later to give a normal form for specifications
akin to Disjunctive Normal Form (\S~\ref{sec:inf-step4}).

@@

\begin{figure}[t]
\vspace{-0.5em}
\begin{align*}
\hspace{-0.5em}
{\small{
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%%
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
%  Q}\trule{\textsc{trans}}
%\qquad
%\\[1.5em]
\vspace{-0.4em}
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
\end{array}}\trule{\textsc{over}} \\[-0.2em]
%\hspace{-0.5em}
%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
%      {R_1 \, \texttt{+} \, R_2 <:^r S_1 \, \texttt{+} \, S_2}
%\trule{\textsc{cong}\texttt{+}}
%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
%      {R_1 \, \texttt{*} \, R_2 <:^r S_1 \, \texttt{*} \, S_2}
%\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
%
\dfrac{R <:^r S}{\term{atLeast}  S <: \term{atLeast}
  R}\trule{\texttt{L}}
\;\;
\dfrac{R <:^r S}{\term{atMost}  R <: \term{atMost}
  S}\trule{\texttt{M}} \\[1.5em]
\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{grow}} \,
\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{shrink}} \,
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}}
 \\[1.5em]
%%
\dfrac{\mathsf{dims}(R) \cap \mathsf{dims}(S) = \emptyset}{R \ast S <: R}
\trule{\textsc{Gen$\ast$}}
\quad
%%
%\\[1.5em]
%\dfrac{R <: S}{\term{readOnce} \, R <: \term{readOnce} \,
%  S}\trule{\textsc{cong}\texttt{R}}
%\;\;
\;
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
%\dfrac{}{R <: \term{atMost} R}
%\quad
%\dfrac{}{R <: \term{atLeast} R}\\[1.5em]
\end{array}}}
\end{align*}
\vspace{-0.5em}
\caption{Approximation $<:^r$ on \textit{regions} and
$<:$ on \textit{specs} (omitting reflexivity, transitivity, and congruences). }
\label{fig:inequations}
\vspace{-0.5em}
\end{figure}

@@

\noindent
We outline here the procedures for checking conformance
of source code against specifications (\S~\ref{subsec:checking})
and for infering specifications from code (\S~\ref{subsec:inference}).
Both rely on a program analysis that converts array indexing
expressions into sets of index schemes. We outline this analysis
first (\S~\ref{subsec:analysis}). Note that the analysis procedure
can be made arbitrarily more complex and wide-ranging independent
of the checking and inference procedures. At the moment, the analysis
is largely \emph{syntactic}, with only a small amount of
semantic interpretation of the code.

\subsection{Analysis of array accesses}
\label{subsec:analysis}

@@

The analysis builds on standard program analyses:
%
\textbf{(1)} basic blocks (CFG);
\textbf{(2)} induction variables per basic block;
\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\textbf{(4)} type information per variable.% where we use the predicate
%\arrayTy{} to classify variables of array type.

The core analysis, captured by the function $\textit{analyse}$ below, then proceeds as follows.
For each assignment statement \textit{assg} within a loop whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed mapping
array variables to \emph{multisets} of indexing schemes:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textit{analyse}(a(\overline{e_1}) = e_2)
 := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2)\}}
\end{array}
\end{align*}
%
The predicate $\neigh(\bar{e_1})$ classifies indices where each
$e \in \bar{e_1}$ is either a
neighbourhood index (based on
the induction variables of the loop, implicit here)
or a constant index.\footnote{If all indices $e \in \bar{e}$ are constants, then
  $\neigh(\bar{e})$ does not hold. Instead $a(\bar{e}$) is
  considered a constant by our analysis, rather
than an array subscript.}% contributing to a stencil.}
Thus, we focus on assignments to an array subscript where the LHS
indices $\overline{e_1}$ are neighbourhood indidces.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a multiset of indexing schemes for its subscripts, computed
with \textsf{schematic} (Def.~\ref{def:schematic}).
Note \textsf{schematic} is undefined ($\bot$) if
$\bar{e}$ contains relative indices which are not neighbourhood
indices. The \textit{analysis} function is used by the
\textit{modelise} function which filters out those arrays
which contain undefined indexing schemes and adds multiplicity
information:
%
\begin{align*}
& \mathit{modelise} (\textit{assg})
= \textit{let} \;\; M = \mathit{analyse}(\textit{assg}) (a) \;\; \textit{in}
\\
& a \mapsto
\begin{cases}
\bot & \bot \in M \\
\textsf{only}(|M|) & \forall u \in M . \, \#(M, u) = 1 \\
\textsf{mult}(|M|) & \forall u \in M . \, \#(M, u) = n
\end{cases}
\end{align*}
where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
multiset $M$ and $|M|$ denotes the set of elements in the multiset
$M$.
Thus, the analysis maps source code to index schemes, which is then
augmented with multiplicity information by $\mathit{modelise}$,
\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.

As a small semantic concession to an otherwise syntactic
analysis: if the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets
then all RHS schemes are \emph{relativised} by
$\textsf{schematic}(\overline{e_1})$. For example,
if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.

\begin{example}
\label{exm:checking}
We demonstrate the analysis and modelise procedures over the
following example program which computes the mean value
of a five-point stencil:
%%
\begin{minted}{fortran}
do i = 1, (n-1)
  do j = 1, (m-1)
    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
    b(i, j) = (a(i, j) + x + y) / 5.0
end do; end do
\end{minted}
%
The \textit{analyse} function matches on
line 4, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
{\small{
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}}}
\end{align*}
The \emph{modelise} function then yields the following
since all multiplicities in the model are one:
%%
\begin{align*}
\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
\end{align*}
%%
\end{example}

@@

Checking verifies the access pattern of a stencil computation in the
source language against its associated specifications. The model of
(\S~\ref{sec:semantics}) guides this process.  Checking
proceeds by generating a model from a specification and generating
a model from the source code, comparing them for
\emph{consistency} base on index scheme
\emph{generalisation} (Def.~\ref{def:generalisation}).

\begin{definition}[Checking]
Let $a$ be an array variable and
$S$ a specification associated to $a$
via \texttt{!= stencil $S$ :: a}.

A stencil kernel
 $v(\overline{e_1}) = e_2$
conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds,
defined:
\begin{align*}
& \textbf{check}_n(S, \textit{assg}, a) =
  \consSub{n}{\textit{modelise}(\textit{assg})(a)}{\interp{S}_n}
&
\end{align*}
Thus, checking applies \textit{modelise}
(\S~\ref{subsec:analysis}) to extract index schemes from source code,
augmented with multiplicitly for the whole set of schemes.
This is compared with the model of the specification $\interp{S}_n$ by
the consistency predicate $\consAName$.
\end{definition}
